---
title: "R data structures"
author: "Colin Rundel"
date: "2018-01-23"
output:
  xaringan::moon_reader:
    css: "slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width=80
)

library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

---
class: middle
count: false

# From last week


---

## Exercise 2

Below is the list of primes between 2 and 100:
```
2, 3, 5, 7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 
43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97
```

If you were given the vector `x = c(3, 4, 12, 19, 23, 48, 50, 61, 63, 78)`, write out the R code necessary to print only the values of `x` that are *not* prime (without using subsetting or the `%in%` operator). 

Your code should use *nested* loops to iterate through the vector of primes and `x`.

---
class: middle
count: false

# Overview

---

## Data structures and dimensionality

.middle[

  Dimensions  |  Homogeneous  | Heterogeneous
--------------|:--------------|:----------------------
  1*d*        | Atomic Vector | List (Generic Vector)
  2*d*        |     Matrix    | Data Frame
  n*d*        |     Array     | `NA`

]

---
class: middle
count: false

# Atomic Vectors

---

## Atomic Vectors

R has six atomic vector types: 

  `typeof`  |  `mode`     |  `storage.mode`
:-----------|:------------|:----------------
logical     |  logical    |  logical
double      |  numeric    |  double
integer     |  numeric    |  integer
character   |  character  |  character
complex     |  complex    |  complex
raw         |  raw        |  raw

---

## Vector types

**logical** - boolean values `TRUE` and `FALSE`

```{r}
typeof(TRUE)
```

**character** - character strings

```{r}
typeof("hello")
typeof('world')
```

---

**double** - floating point numerical values (default numerical type)

```{r}
typeof(1.33)
typeof(7)
```

**integer** - integer numerical values (indicated with an `L`)

```{r}
typeof( 7L )
typeof( 1:3 )
```

---

## Concatenation

Atomic vectors can be constructed using the concatenate, `c()`, function, *note* that vectors are always flat.

```{r}
c(1,2,3)
c("Hello", "World!")
c(1,c(2, c(3)))
```

---
class: split-thirds

## Testing types

* `typeof(x)` - returns a character vector (length 1) of the *type* of object `x`.

* `mode(x)` - returns a character vector (length 1) of the *mode* of object `x`.

* `storage.mode(x)` - returns a character vector (length 1) of the *mode* of object `x`.

.column[
```{r}
typeof(1)
typeof(1L)
typeof("A")
typeof(TRUE)
```
]

.column[
```{r}
mode(1)
mode(1L)
mode("A")
mode(TRUE)
```
]

.column[
```{r}
storage.mode(1)
storage.mode(1L)
storage.mode("A")
storage.mode(TRUE)
```
]


---
class: split-thirds

## Logical Predicates

* `is.logical(x)` - returns `TRUE` if `x` has *type* logical.

* `is.character(x)` - returns `TRUE` if `x` has *type* character.

* `is.double(x)` - returns `TRUE` if `x` has *type* double.

* `is.integer(x)` - returns `TRUE` if `x` has *type* integer.

* `is.numeric(x)` - returns `TRUE` if `x` has *mode* numeric.

.column[
```{r}
is.integer(1)
is.integer(1L)
is.integer(3:7)
```
]

.column[
```{r}
is.double(1)
is.double(1L)
is.double(3:8)
```
]

.column[
```{r}
is.numeric(1)
is.numeric(1L)
is.numeric(3:7)
```
]


---

## Other useful predicates

* `is.atomic(x)` - returns `TRUE` if `x` is an *atomic vector*.

* `is.vector(x)` - returns `TRUE` if `x` is either type of vector (i.e. either *atomic vector* or *list*).

<br/>

```{r}
is.atomic(c(1,2,3))
is.vector(c(1,2,3))
is.atomic(list(1,2,3))
is.vector(list(1,2,3))
```


---

## Type Coercion

R is a dynamically typed language -- it will automatically convert between most type without raising warnings or errors.

```{r}
c(1,"Hello")
c(FALSE, 3L)
c(1.2, 3L)
```

---

## Operator coercion

Functions and operators will attempt to coerce object to an appropriate type

```{r}
3.1+1L
log(TRUE)
TRUE & 7
FALSE | !5
```


---
class: split-50

## Explicit Coercion

Most of the `is` functions we just saw have an `as` variant which can be used for *explicit* coercion.

.column[
```{r}
as.logical(5.2)
as.character(TRUE)
as.integer(pi)
```
]

.column[
```{r}
as.numeric(FALSE)
as.double("7.2")
as.double("one")
```
]


---

## Missing Values

R uses `NA` to represent missing values in its data structures, what may not be obvious is that there are different `NA`s for the different types.

```{r}
typeof(NA)
typeof(NA+1)
typeof(NA+1L)
```


---
class: split-50

## Contagiousness of Missing Values

Because `NA`s represent missing values it makes sense that any calculation using them should also be missing.

.column[
```{r}
1 + NA
1 / NA
NA * 5
```
]

.column[
```{r}
mean(c(1,2,3,NA))
sqrt(NA)
3^NA
```
]

---

## Conditionals and missing values

`NA`s can be problematic in some cases (particularly for control flow)

```{r error=TRUE}
1 == NA

if (2 != NA)
  "Here"

if (any(c(1,2,NA,4) > 2))
  "There"
```


---
class: split-50

## Testing for `NA`

To explicitly test if a value is missing it is necessary to use `is.na` (often along with `any` or `all`).

.column[
```{r}
is.na(NA)
is.na(1)
is.na(c(1,2,3,NA))
```
]

.column[
```{r}
any(is.na(c(1,2,3,NA)))
all(is.na(c(1,2,3,NA)))
```
]


---
class:split-50

## Other Special values

* `NaN` - Not a number

* `Inf` - Positive infinity

* `-Inf` - Negative infinity



.column[
```{r}
pi / 0
0 / 0
1/0 + 1/0
```
]

.column[
```{r}
1/0 - 1/0
NaN / NA
NaN * NA
```
]


---

## Testing for `inf` and `NaN`

`NaN` and `Inf` don't have the same testing issues that `NA` has, but there are still convenience functions for testing for 

```{r}
is.finite(-Inf)
is.finite(1/0+1/0)
is.nan(1/0-1/0)
```


---
class: split-50

## Coercion for infinity and NaN

First remember that `Inf`, `-Inf`, and `NaN` have type double, however their coercion behavior is not the same as for other double values.

```{r}
as.integer(Inf)
as.integer(NaN)
```

.column[
```{r}
as.logical(Inf)
as.logical(NaN)
```
]

.column[
```{r}
as.character(Inf)
as.character(NaN)
```
]

---

## Exercise 1

**Part 1**

What is the type of the following vectors? Explain why they have that type.

* `c(1, NA+1L, "C")`
* `c(1L / 0, NA)`
* `c(1:3, 5)`
* `c(3L, NaN+1L)`
* `c(NA, TRUE)`


**Part 2**

Considering only the four (common) data types, what is R's implicit type conversion hierarchy (from highest priority to lowest priority)? 

Hint - think about the pairwise interactions between types.

---
class: middle
count: false

# Generic Vectors


---

## Lists

Lists are _generic vectors_, in that they are 1d and can contain any combination of R objects.

```{r}
list("A", c(TRUE,FALSE), (1:4)/2, function(x) x^2)
```


---

## Structure

Often we want a more compact representation of a complex object, the `str` function is useful for this particular task

```{r}
str( list("A", c(TRUE,FALSE), (1:4)/2, function(x) x^2) )
```


---

## Recursive lists

Lists can even contain other lists, meaning they don't have to be flat

```{r}
str( list(1, list(2, list(3, 4), 5)) )
```


---

## List Coercion

By default a vector will be coerced to a list (as a list is more generic) if needed

```{r}
str( c(1, 2, list(4, 5, list(6, 7))) )
```

--

We can force a list into an atomic vector using `unlist`

```{r}
unlist(list(1:3, list(4:5, 6)))
unlist( list(1, list(2, list(3, "Hello"))) )
```


---

## Named lists

Because of their more complex structure we often want to name the elements of a list (we can also do this with vectors). This can make reading and accessing the list more straight forward.

```{r}
str(list(A = 1, B = list(C = 2, D = 3)))
list("knock knock" = "who's there?")
names(list(ABC=1, DEF=list(H=2, I=3)))
```


---

## Exercise 2

Represent the following JSON data as a list in R.

```json
{
  "firstName": "John",
  "lastName": "Smith",
  "age": 25,
  "address": 
  {
    "streetAddress": "21 2nd Street",
    "city": "New York",
    "state": "NY",
    "postalCode": 10021
  },
  "phoneNumber": 
  [
    {
      "type": "home",
      "number": "212 555-1239"
    },
    {
      "type": "fax",
      "number": "646 555-4567"
    }
  ]
}
```


---

# Acknowledgments
## Acknowledgments

Above materials are derived in part from the following sources:

* Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)
* [R Language Definition](http://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html)
